---
title: "Occupancy and N-mixture Models for Gilded Sapphire Hummingbirds"
author: "Facundo X. Palacio"
date: "June 11, 2025"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
---

```{r setup, include=TRUE}
# Global knitr options for cleaner output and consistency
knitr::opts_chunk$set(
  echo = TRUE,      # Display code in output
  message = FALSE,  # Suppress messages from packages/functions
  warning = FALSE,  # Suppress warnings
  cache = FALSE,    # Do not cache chunks (set to TRUE for long-running chunks once stable)
  fig.align = "center", # Center figures by default
  fig.width = 7,    # Default figure width
  fig.height = 5    # Default figure height
)
```

# Introduction

This R Markdown document presents an analysis using Occupancy and N-mixture models for Gilded Sapphire hummingbirds. The analysis is divided into two main sections: Occupancy Models using `unmarked` package and N-mixture Models using `R2jags`.

The R script provided loads detection history and covariate data, prepares it for modeling, fits several models, compares them, and generates plots for the best-fitting model.

# 1. Occupancy Models

This section focuses on multi-season occupancy modeling using the `unmarked` package. We will load and prepare the data, build an `unmarkedMultFrame` object, fit several `colext` models, and compare them using AICc.

## 1.1 Load packages
```{r load-libraries-occupancy}
library(unmarked)
library(AICcmodavg)
library(tidyr)
library(dplyr)
```

## 1.2 Load and prepare data

Detection history, observation covariates (hour and flower abundance), site covariates (habitat type), and yearly site covariates (mean flower abundance per season) are loaded and formatted.
```{r load-data-occupancy}
# Load detection history (10 sites with 12 visits each)
humm_abundance <- read.table("../Data/data_hummingbirds_lp.txt", head = T)
detection_history <- 1*(humm_abundance[, 2:13]>0)

# Split detection history by season
spring <- data.frame(detection_history[, c(1:2, 12)], y1.na = NA)
summer <- data.frame(detection_history[, 3:5], y2.na = NA)
fall <- detection_history[, 6:9]
winter <- data.frame(detection_history[, 10:11], y3.na = NA, y4.na = NA)
y <- cbind(spring, summer, fall, winter)

# Mean detection across seasons (for descriptive purposes)
cat("Mean detection in Spring:", mean(as.matrix(spring[,1:3])), "\n")
cat("Mean detection in Summer:", mean(as.matrix(summer[,1:3])), "\n")
cat("Mean detection in Fall:", mean(as.matrix(fall[,1:4])), "\n")
cat("Mean detection in Winter:", mean(as.matrix(winter[,1:2])), "\n")

# Load observation covariates
ObsCovs <- read.table("../Data/obs_covs_hummingbirds_lp.txt", head = T)

# Reshape hour covariate to wide format for each season
hour_wide <- ObsCovs %>%
  select(ID_transecta, visit, hora_decimal) %>%
  pivot_wider(
    names_from = visit,
    values_from = hora_decimal
  )
hour_spring <- data.frame(hour_wide[, c(2:3, 13)], x1.na = NA)
hour_summer <- data.frame(hour_wide[, 4:6], x2.na = NA)
hour_fall <- hour_wide[, 7:10]
hour_winter <- data.frame(hour_wide[, 11:12], x3.na = NA, x4.na = NA)
hour_wide <- cbind(hour_spring, hour_summer, hour_fall, hour_winter)

# Reshape flower abundance covariate to wide format for each season
flower_wide <- ObsCovs %>%
  select(ID_transecta, visit, flower_abun) %>%
  pivot_wider(
    names_from = visit,
    values_from = flower_abun
  )
flower_spring <- data.frame(flower_wide[, c(2:3, 13)], x1.na = NA)
flower_summer <- data.frame(flower_wide[, 4:6], x2.na = NA)
flower_fall <- flower_wide[, 7:10]
flower_winter <- data.frame(flower_wide[, 11:12], x3.na = NA, x4.na = NA)
flower_wide <- cbind(flower_spring, flower_summer, flower_fall, flower_winter)

# Load site covariates
siteCovs <- read.table("../Data/site_covs_hummingbirds_lp.txt", head = T)
siteCovs$habitat_type <- as.factor(siteCovs$habitat_type)

# Load yearly site covariates
yearlySiteCovs <- read.table("../Data/yearly_site_covs_hummingbirds_lp.txt", head = T)
flower_matrix <- yearlySiteCovs %>%
  arrange(ID_transecta, season) %>%
  pivot_wider(names_from = season,
              values_from = mean_flower_abundance,
              names_prefix = "season") %>%
  select(starts_with("season")) %>%
  as.matrix()
flower_matrix <- flower_matrix[, c(2:3,1,4)] # ensure order (Spring, Summer, Fall, Winter)

# Check dimensions of prepared data
dim(y)
dim(flower_wide)
dim(hour_wide)
dim(siteCovs)
dim(flower_matrix)
```

## 1.3 Build `unmarkedMultFrame` Object

The `unmarkedMultFrame` object is created, which is a specialized data structure required by `unmarked` for multi-season occupancy models.

```{r build-unmarkedmultframe}
humm_unmarkedMultFrame <- unmarkedMultFrame(
  y = as.matrix(y),
  numPrimary = 4, # 4 primary seasons (spring, summer, fall, winter)
  obsCovs = list(hour = hour_wide,
                 flower_abun = flower_wide),
  siteCovs = data.frame(habitat_type = siteCovs[,-1]), # Exclude ID column
  yearlySiteCovs = list(mean_flower_abundance = flower_matrix)
)
```

## 1.4 Multi-season Occupancy Models

Several `colext` models are fitted to analyze occupancy, colonization, extinction, and detection probabilities under different covariate structures.

### Null Model (`fm0`)

This model assumes constant occupancy, colonization, extinction, and detection probabilities.

```{r fit-fm0}
fm0 <- colext(
  psiformula = ~1,
  gammaformula = ~1,
  epsilonformula = ~1,
  pformula = ~1,
  data = humm_unmarkedMultFrame,
  method = "BFGS"
)
print(summary(fm0))
```

### Time Model (`fm1`)

This model investigates the effect of `hour` and `flower_abun` on detection probability, while other parameters remain constant.

```{r fit-fm1}
fm1 <- colext(
  psiformula = ~1,
  gammaformula = ~1,
  epsilonformula = ~1,
  pformula = ~hour + flower_abun,
  data = humm_unmarkedMultFrame,
  method = "BFGS"
)
print(summary(fm1))
```

### Habitat Model (`fm2`)

This model explores the influence of `habitat_type` on colonization and extinction probabilities, with detection depending on `hour`.

```{r fit-fm2}
fm2 <- colext(
  psiformula = ~1,
  gammaformula = ~habitat_type,
  epsilonformula = ~habitat_type,
  pformula = ~hour,
  data = humm_unmarkedMultFrame,
  method = "BFGS"
)
print(summary(fm2))
```

### Flower Model (`fm3`)

This model assesses the impact of mean `flower_abundance` on colonization and extinction probabilities, with detection depending on `hour` and `flower_abun`.

```{r fit-fm3}
fm3 <- colext(
  psiformula = ~1,
  gammaformula = ~mean_flower_abundance, # Note: using the yearlySiteCovs for gamma/epsilon
  epsilonformula = ~mean_flower_abundance,
  pformula = ~hour + flower_abun,
  data = humm_unmarkedMultFrame,
  method = "BFGS"
)
print(summary(fm3))
```

## 1.5 Model Comparison (AICc)

The models are compared using AICc to determine the best-fitting model.

```{r aic-table-occupancy}
models <- list(
  'p(.)psi(.)' = fm0,
  'p(hour, flower abundance)' = fm1,
  'p(hour)gamma(habitat_type)epsilon(habitat_type)' = fm2,
  'p(hour, flower_abundance)gamma(mean_flower_abundance)epsilon(mean_flower_abundance)' = fm3
)

aicTable <- aictab(models)
print(aicTable)
# write.csv(aicTable, "aicTable.csv", row.names = FALSE)
```

## 1.6 Back-transformed Parameters (Null Model Example)

Example of back-transforming parameters for the null model to get probabilities/rates.

```{r backtransform-fm0}
backTransform(fm0, type = "psi")
backTransform(fm0, type = "col")
backTransform(fm0, type = "ext")
backTransform(fm0, type = "det")
```

# 2. N-mixture Models

This section delves into N-mixture models, which estimate abundance while accounting for imperfect detection. We will use `R2jags` for Bayesian inference.

## 2.1 Load packages

```{r load-libraries-nmixture}
library(R2jags)
library(coda)
library(dplyr)
library(tidyr)
```

## 2.2 Reformat Data for N-mixture Models

The data needs to be restructured into a 3D array (`y_array`, `hour_array`, `flower_array`) suitable for JAGS, with dimensions [sites, seasons, surveys].

```{r reformat-data-nmixture}
humm_abundance <- read.table("../Data/data_hummingbirds_lp.txt", head = T)
y_wide <- humm_abundance[, 2:13]
Jmax <- 4  # max surveys per season
T_seasons <- 4     # number of primary periods (seasons)
I_sites <- nrow(y_wide)

spring <- data.frame(y_wide[, c(1:2, 12)], y1.na = NA)
summer <- data.frame(y_wide[, 3:5], y2.na = NA)
fall <- y_wide[, 6:9]
winter <- data.frame(y_wide[, 10:11], y3.na = NA, y4.na = NA)
y <- cbind(spring, summer, fall, winter)

# Load covariate data (already loaded for occupancy, but re-run for clarity)
ObsCovs <- read.table("../Data/obs_covs_hummingbirds_lp.txt", head = T)

hour_wide <- ObsCovs %>%
  select(ID_transecta, visit, hora_decimal) %>%
  pivot_wider(
    names_from = visit,
    values_from = hora_decimal
  )
hour_spring <- data.frame(hour_wide[, c(2:3, 13)], x1.na = NA)
hour_summer <- data.frame(hour_wide[, 4:6], x2.na = NA)
hour_fall <- hour_wide[, 7:10]
hour_winter <- data.frame(hour_wide[, 11:12], x3.na = NA, x4.na = NA)
hour_wide <- cbind(hour_spring, hour_summer, hour_fall, hour_winter)

flower_wide <- ObsCovs %>%
  select(ID_transecta, visit, flower_abun) %>%
  pivot_wider(
    names_from = visit,
    values_from = flower_abun
  )
flower_spring <- data.frame(flower_wide[, c(2:3, 13)], x1.na = NA)
flower_summer <- data.frame(flower_wide[, 4:6], x2.na = NA)
flower_fall <- flower_wide[, 7:10]
flower_winter <- data.frame(flower_wide[, 11:12], x3.na = NA, x4.na = NA)
flower_wide <- cbind(flower_spring, flower_summer, flower_fall, flower_winter)

siteCovs <- read.table("../Data/site_covs_hummingbirds_lp.txt", head = T)
siteCovs$habitat_type <- as.factor(siteCovs$habitat_type)

# Function to split wide matrices into 3D arrays
split_by_season <- function(mat_wide, T, Jmax) {
  array_3d <- array(NA, dim = c(nrow(mat_wide), T, Jmax))
  for (t in 1:T) {
    cols <- ((t - 1)*Jmax + 1):(t*Jmax)
    array_3d[, t, ] <- as.matrix(mat_wide[, cols])
  }
  return(array_3d)
}

y_array <- split_by_season(y, T = T_seasons, Jmax = Jmax)
hour_array <- split_by_season(hour_wide, T = T_seasons, Jmax = Jmax)
hour_array[is.na(hour_array)] <- 1 # Replace NA with 1 (or appropriate default)
flower_array <- split_by_season(flower_wide, T = T_seasons, Jmax = Jmax)
flower_array[is.na(flower_array)] <- 1 # Replace NA with 1 (or appropriate default)

# Site-level covariates
habitat_type <- as.numeric(siteCovs$habitat_type) # Convert to numeric for JAGS

# Yearly site-level covariates (flower abundance per season)
yearlySiteCovs <- read.table("../Data/yearly_site_covs_hummingbirds_lp.txt", head = T)
flower_matrix <- yearlySiteCovs %>%
  arrange(ID_transecta, season) %>%
  pivot_wider(names_from = season,
              values_from = mean_flower_abundance,
              names_prefix = "season") %>%
  select(starts_with("season")) %>%
  as.matrix()
flower_matrix <- flower_matrix[, c(2:3,1,4)] # ensure order

# Define constants for JAGS models
n_sites <- dim(y_array)[1]
n_seasons <- dim(y_array)[2]
n_surveys <- dim(y_array)[3]
```

## 2.3 N-mixture Model: Null Model (`fm0`)

This is a basic N-mixture model without any covariates, modeling initial abundance, survival, recruitment, and detection as constants.

```{r nmixture-fm0}
# Model definition for JAGS
fm0_jags_code <- "
model {
  # Priors
  lambda ~ dunif(0, 20)      # Initial abundance (Poisson mean)
  phi ~ dunif(0, 1)         # Apparent survival probability
  gamma ~ dunif(0, 20)      # Recruitment/colonization (Poisson mean)
  p ~ dunif(0, 1)           # Detection probability

  for (i in 1:n_sites) {
    N[i,1] ~ dpois(lambda) # Initial abundance at each site

    for (t in 2:n_seasons) {
      S[i,t-1] ~ dbin(phi, N[i,t-1]) # Number of survivors from N[i,t-1]
      R[i,t] ~ dpois(gamma)          # Number of recruits
      N[i,t] <- S[i,t-1] + R[i,t]    # Abundance at current season
    }

    for (t in 1:n_seasons) {
      for (j in 1:max_surveys) {
        y[i,t,j] ~ dbin(p, N[i,t]) # Observed counts are binomial draws from true abundance
      }
    }
  }
}
"

# Data list for JAGS
data_jags_fm0 <- list(
  y = y_array,
  n_sites = n_sites,
  n_seasons = n_seasons,
  max_surveys = 4 # Max surveys per season
)

# Initial values for latent states (N) and parameters
inits_fm0 <- function() {
  N_init <- matrix(NA, nrow = n_sites, ncol = n_seasons)
  # Set initials for season 1 based on max observed counts
  N_init[,1] <- apply(y_array[,1,,drop=FALSE], 1, max, na.rm=TRUE) + 10
  # For subsequent seasons, N can be NA as JAGS will initialize them
  for (t in 2:n_seasons) {
    N_init[,t] <- NA
  }

  list(
    lambda = runif(1, 1, 5),
    phi = runif(1, 0.5, 1),
    gamma = runif(1, 0, 3),
    p = runif(1, 0.1, 0.9),
    N = N_init
  )
}

# Parameters to monitor
params_fm0 <- c("lambda", "phi", "gamma", "p", "N")

# Run the JAGS model (set eval=TRUE to avoid running long computation during knitting)
set.seed(99)
fit_fm0 <- jags(
  data = data_jags_fm0,
  inits = inits_fm0,
  parameters.to.save = params_fm0,
  model.file = textConnection(fm0_jags_code),
  n.chains = 3,
  n.iter = 100000,
  n.burnin = 10000,
  n.thin = 50
)

# Print and summarize results
print(fit_fm0)
rhat_values <- fit_fm0$BUGSoutput$summary[, "Rhat"]
cat("Summary of Rhat values for fm0:\n")
print(summary(rhat_values))

# Export summary to CSV
# summary_fm0 <- as.data.frame(fit_fm0$BUGSoutput$summary)
# write.csv(summary_fm0, "fit_fm0_summary.csv", row.names = TRUE)

# Save posterior samples (and load if needed later)
# saveRDS(fit_fm0, file = "posterior_samples_fm0.rds")
# fit_fm0 <- readRDS("posterior_samples_fm0.rds")
```

## 2.4 N-mixture Model: Time Model (`fm1`)

This model includes covariates for `hour` and `flower_abun` affecting detection probability.

```{r nmixture-fm1}
# Model definition for JAGS
fm1_jags_code <- "
  model {
  # Priors
  lambda ~ dunif(0, 20)        # Initial abundance
  phi ~ dunif(0, 1)           # Survival probability
  gamma ~ dunif(0, 20)        # Recruitment rate

  beta0 ~ dnorm(0, 0.001)        # Intercept for detection on logit scale
  beta_hour ~ dnorm(0, 0.001)   # Effect of hour on detection
  beta_flower_p ~ dnorm(0, 0.01) # Effect of flower abundance on detection

  for (i in 1:n_sites) {
    N[i,1] ~ dpois(lambda) # Initial abundance

    for (t in 2:n_seasons) {
      S[i,t-1] ~ dbin(phi, N[i,t-1]) # Number of survivors
      R[i,t] ~ dpois(gamma)          # Number of recruits
      N[i,t] <- S[i,t-1] + R[i,t]    # Abundance in current season
    }

    for (t in 1:n_seasons) {
      for (j in 1:max_surveys) {
        # Logit-linear model for detection probability
        logit(p[i,t,j]) <- beta0 + beta_hour*hour[i,t,j] + beta_flower_p*flower_abun[i,t,j]
        y[i,t,j] ~ dbin(p[i,t,j], N[i,t]) # Observed counts
      }
    }
  }
}
"
# Data list for JAGS
data_jags_fm1 <- list(
  y = y_array,
  hour = hour_array,
  n_sites = n_sites,
  n_seasons = n_seasons,
  max_surveys = 4,
  flower_abun = flower_array
)

# Initial values for latent states (N) and parameters
inits_fm1 <- function() {
  N_init <- matrix(NA, nrow = n_sites, ncol = n_seasons)
  for (i in 1:n_sites) {
    max_y_all <- max(y_array[i,,], na.rm = TRUE)
    N_init[i, 1] <- ifelse(is.finite(max_y_all), max_y_all + 50, 50) # Initialize N[i,1]
  }

  list(
    lambda = runif(1, 1, 5),
    phi = runif(1, 0.5, 1),
    gamma = runif(1, 0, 3),
    beta0 = rnorm(1, 0, 1),
    beta_hour = rnorm(1, 0, 1),
    beta_flower_p = rnorm(1, 0, 1),
    N = N_init
  )
}

# Parameters to monitor
params_fm1 <- c("lambda", "phi", "gamma", "beta0",
                "beta_hour", "beta_flower_p", "N")

# Run the JAGS model (set eval=TRUE to avoid running long computation during knitting)
set.seed(99)
fit_fm1 <- jags(
  data = data_jags_fm1,
  inits = inits_fm1,
  parameters.to.save = params_fm1,
  model.file = textConnection(fm1_jags_code),
  n.chains = 6,
  n.iter = 100000,
  n.burnin = 10000,
  n.thin = 50
)

print(fit_fm1)

# Traceplot and summary of Rhat values for convergence diagnostics
# fit_fm1 <- readRDS("posterior_samples_fm1.rds") # Load if not run above
traceplot(fit_fm1)

mcmc_samples <- as.mcmc(fit_fm1)
mean_deviance_per_chain <- sapply(mcmc_samples, function(chain) {
  mean(chain[, "deviance"])
})
cat("Mean deviance per chain:\n")
print(mean_deviance_per_chain)

# Subsetting chains (e.g., if some chains did not converge well)
subset_chains <- mcmc_samples[4:6]

# Compute summary statistics and Gelman-Rubin diagnostics for selected chains
summary_stats <- summary(subset_chains)
rhat_diag <- gelman.diag(subset_chains, multivariate = FALSE)$psrf

# Combine results into a data frame
fit_fm1_summary <- data.frame(
  Parameter = rownames(summary_stats$statistics),
  Mean = summary_stats$statistics[, "Mean"],
  SD = summary_stats$statistics[, "SD"],
  `2.5%` = summary_stats$quantiles[, "2.5%"],
  `97.5%` = summary_stats$quantiles[, "97.5%"],
  Rhat = rhat_diag[, "Point est."],
  Rhat_UpperCI = rhat_diag[, "Upper C.I."]
)

print(fit_fm1_summary)

# Export summary to CSV
# write.csv(fit_fm1_summary, "fit_fm1_summary.csv", row.names = TRUE)

# Save posterior samples
# saveRDS(fit_fm1, file = "posterior_samples_fm1.rds")
```

## 2.5 N-mixture Model: Habitat Model (`fm2`)

This model explores the effect of `habitat_type` on colonization and extinction, with `hour` affecting detection.

```{r nmixture-fm2}
# Model definition for JAGS
fm2_jags_code <- "
model {
  # Priors
  lambda ~ dgamma(1, 0.1)           # Mean initial abundance

  alpha_p ~ dnorm(0, 0.1)           # Intercept for detection
  beta_hour ~ dnorm(0, 0.1)         # Effect of hour on detection

  alpha_gamma ~ dnorm(0, 0.1)       # Intercept for colonization (logit scale)
  beta_gamma ~ dnorm(0, 0.1)        # Effect of habitat_type on colonization
  # Using habitat[i] as a numeric value. If habitat_type is a factor with >2 levels,
  # it should be handled with multiple beta coefficients (e.g., beta_gamma[habitat[i]])

  alpha_phi ~ dnorm(0, 0.1)     # Intercept for survival (logit scale)
  beta_phi ~ dnorm(0, 0.1)      # Effect of habitat_type on survival (phi = 1 - extinction)

  for (i in 1:n_sites) {
    # Initial abundance
    N[i,1] ~ dpois(lambda)

    # Detection process for season 1
    for (j in 1:n_visits) {
      logit(p[i,1,j]) <- alpha_p + beta_hour*hour[i,1,j]
      y[i,1,j] ~ dbin(p[i,1,j], N[i,1])
    }

    for (t in 2:n_seasons) {
      # Colonization and Extinction (on logit scale)
      logit(gamma[i,t-1]) <- alpha_gamma + beta_gamma*habitat[i]
      logit(phi[i,t-1]) <- alpha_phi + beta_phi*habitat[i]

      S[i,t-1] ~ dbin(phi[i,t-1], N[i,t-1])  # Number of survivors (phi is survival)
      G[i,t-1] ~ dpois(gamma[i,t-1])         # Number of colonizers (recruits)
      N[i,t] <- S[i,t-1] + G[i,t-1]          # Abundance at current season

      # Detection process for subsequent seasons
      for (j in 1:n_visits) {
        y[i,t,j] ~ dbin(p[i,t,j], N[i,t])
        logit(p[i,t,j]) <- alpha_p + beta_hour*hour[i,t,j]
      }
    }
  }
}
"
# Prepare data for JAGS
jags_data_fm2 <- list(
  n_sites = dim(y_array)[1],
  n_seasons = dim(y_array)[2],
  n_visits = dim(y_array)[3],
  y = y_array,
  hour = hour_array,
  habitat = as.numeric(siteCovs$habitat_type) # Ensure habitat is numeric (1, 2, ...)
)

# Initial values
inits_fm2 <- function() {
  N_init <- matrix(NA, nrow = jags_data_fm2$n_sites, ncol = jags_data_fm2$n_seasons)
  for (i in 1:jags_data_fm2$n_sites) {
    max_y1 <- max(jags_data_fm2$y[i, 1, ], na.rm = TRUE)
    N_init[i, 1] <- ifelse(is.finite(max_y1), max_y1 + 300, 300)
  }

  list(
    alpha_gamma = rnorm(1, 0, 1),
    beta_gamma = rnorm(1, 0, 1),
    alpha_phi = rnorm(1, 0 , 1),
    beta_phi = rnorm(1, 0, 1),
    alpha_p = rnorm(1, 0, 1),
    beta_hour = rnorm(1, 0, 1),
    N = N_init
  )
}

# Parameters to monitor
params_fm2 <- c("lambda",
            "alpha_gamma", "beta_gamma",
            "alpha_phi", "beta_phi",
            "alpha_p", "beta_hour")

# Run JAGS model (set eval=TRUE to avoid running long computation during knitting)
set.seed(99)
fit_fm2 <- jags(
  data = jags_data_fm2,
  inits = inits_fm2,
  parameters.to.save = params_fm2,
  model.file = textConnection(fm2_jags_code),
  n.chains = 3,
  n.iter = 100000,
  n.burnin = 50000,
  n.thin = 50
)

# Print and summarize results
rhat_values <- fit_fm2$BUGSoutput$summary[, "Rhat"]
cat("Summary of Rhat values for fm2:\n")
print(summary(rhat_values))
print(fit_fm2)
traceplot(fit_fm2)

# Export summary to CSV
# summary_fm2 <- as.data.frame(fit_fm2$BUGSoutput$summary)
# write.csv(summary_fm2, "fit_fm2_summary.csv", row.names = TRUE)

# Save posterior samples
# saveRDS(fit_fm2, file = "posterior_samples_fm2.rds")
# fit_fm2 <- readRDS("posterior_samples_fm2.rds") # Load if not run above
```

## 2.6 N-mixture Model: Flower Model (`fm3`)

This model incorporates the effect of mean `flower_abundance` on colonization and extinction, and individual `flower_abun` on detection.

```{r nmixture-fm3}
# Model definition for JAGS
fm3_jags_code <- "
model {
  # Priors
  lambda ~ dgamma(1, 0.1)             # Mean initial abundance

  alpha_p ~ dnorm(0, 0.1)             # Intercept for detection
  beta_hour ~ dnorm(0, 0.1)           # Effect of hour on detection
  beta_flower_p ~ dnorm(0, 0.1)       # Effect of flower abundance on detection (observation level)

  alpha_gamma ~ dnorm(0, 0.1)         # Intercept for colonization (logit)
  beta_gamma ~ dnorm(0, 0.1)          # Effect of mean flower abundance on colonization (yearly level)

  alpha_phi ~ dnorm(0, 0.1)           # Intercept for survival (logit)
  beta_phi ~ dnorm(0, 0.1)            # Effect of mean flower abundance on survival (yearly level)

  for (i in 1:n_sites) {
    N[i,1] ~ dpois(lambda) # Initial abundance

    for (j in 1:n_visits) {
      logit(p[i,1,j]) <- alpha_p + beta_hour*hour[i,1,j] + beta_flower_p*flower_abun[i,1,j]
      y[i,1,j] ~ dbin(p[i,1,j], N[i,1])
    }

    for (t in 2:n_seasons) {
      # Colonization and Survival (phi = 1 - extinction)
      logit(gamma[i,t-1]) <- alpha_gamma + beta_gamma*mean_flower[i,t-1]
      logit(phi[i,t-1]) <- alpha_phi + beta_phi*mean_flower[i,t-1]

      S[i,t-1] ~ dbin(phi[i,t-1], N[i,t-1]) # Number of survivors
      G[i,t-1] ~ dpois(gamma[i,t-1])        # Number of recruits
      N[i,t] <- S[i,t-1] + G[i,t-1]         # Abundance at current season

      for (j in 1:n_visits) {
        logit(p[i,t,j]) <- alpha_p + beta_hour*hour[i,t,j] + beta_flower_p*flower_abun[i,t,j]
        y[i,t,j] ~ dbin(p[i,t,j], N[i,t])
      }
    }
  }
}
"

jags_data_fm3 <- list(
  n_sites = dim(y_array)[1],
  n_seasons = dim(y_array)[2],
  n_visits = dim(y_array)[3],
  y = y_array,
  hour = hour_array,
  flower_abun = flower_array,
  mean_flower = flower_matrix[,2:4] # Seasons 2, 3, 4 for colonization/extinction from seasons 1, 2, 3
)

inits_fm3 <- function() {
  N_init <- matrix(NA, nrow = jags_data_fm3$n_sites, ncol = jags_data_fm3$n_seasons)
  for (i in 1:jags_data_fm3$n_sites) {
    max_y1 <- max(jags_data_fm3$y[i, 1, ], na.rm = TRUE)
    N_init[i, 1] <- ifelse(is.finite(max_y1), max_y1 + 800, 800)
  }

  list(
    alpha_gamma = rnorm(1, 0, 1),
    beta_gamma = rnorm(1, 0, 1),
    alpha_phi = rnorm(1, 0, 1),
    beta_phi = rnorm(1, 0, 1),
    alpha_p = rnorm(1, 0, 1),
    beta_hour = rnorm(1, 0, 1),
    beta_flower_p = rnorm(1, 0, 1),
    N = N_init
  )
}

params_fm3 <- c("lambda", "alpha_gamma", "beta_gamma",
  "alpha_phi", "beta_phi", "alpha_p", "beta_hour", "beta_flower_p")

set.seed(10)
fit_fm3 <- jags(
  data = jags_data_fm3,
  inits = inits_fm3,
  parameters.to.save = params_fm3,
  model.file = textConnection(fm3_jags_code),
  n.chains = 6,
  n.iter = 100000,
  n.burnin = 50000,
  n.thin = 50
)

# Traceplot and summary of Rhat values for convergence diagnostics
traceplot(fit_fm3)
mcmc_samples <- as.mcmc(fit_fm3)
mean_deviance_per_chain <- sapply(mcmc_samples, function(chain) {
  mean(chain[, "deviance"])
})
cat("Mean deviance per chain:\n")
print(mean_deviance_per_chain)

# Subsetting chains (e.g., if some chains did not converge well)
subset_chains <- mcmc_samples[2:4]

# Compute summary statistics and Gelman-Rubin diagnostics for selected chains
summary_stats <- summary(subset_chains)
rhat_diag <- gelman.diag(subset_chains, multivariate = FALSE)$psrf

# Combine results into a data frame
fit_fm3_summary <- data.frame(
  Parameter = rownames(summary_stats$statistics),
  Mean = summary_stats$statistics[, "Mean"],
  SD = summary_stats$statistics[, "SD"],
  `2.5%` = summary_stats$quantiles[, "2.5%"],
  `97.5%` = summary_stats$quantiles[, "97.5%"],
  Rhat = rhat_diag[, "Point est."],
  Rhat_UpperCI = rhat_diag[, "Upper C.I."]
)

print(fit_fm3_summary)

# Export summary to CSV
# write.csv(fit_fm3_summary, "fit_fm3_summary.csv", row.names = TRUE)

# Save posterior samples
# saveRDS(fit_fm3, file = "posterior_samples_fm3.rds")
# fit_fm3 <- readRDS("posterior_samples_fm3.rds") # Load if not run above
```

## 2.7 Model Comparisons (DIC)

Compare the N-mixture models using Deviance Information Criterion (DIC).

```{r dic-comparison}
# Print DIC and pV for each model
cat("DIC for fm0:", fit_fm0$BUGSoutput$DIC, "\tpV for fm0:", fit_fm0$BUGSoutput$pV, "\n")
cat("DIC for fm1:", fit_fm1$BUGSoutput$DIC, "\tpV for fm1:", fit_fm1$BUGSoutput$pV, "\n")
cat("DIC for fm2:", fit_fm2$BUGSoutput$DIC, "\tpV for fm2:", fit_fm2$BUGSoutput$pV, "\n")
cat("DIC for fm3:", fit_fm3$BUGSoutput$DIC, "\tpV for fm3:", fit_fm3$BUGSoutput$pV, "\n")
```

# 3. Best Model Plots (from `fm1`)

This section generates plots based on the N-mixture `fm1` model (Time Model), which includes effects of `hour` and `flower_abun` on detection. It visualizes the estimated detection probability as a function of these covariates and also seasonal abundance.

```{r best-model-plots}
# Load the best model's posterior samples
#fit_fm1 <- readRDS("posterior_samples_fm1.rds")
#ObsCovs <- read.table("../Data/obs_covs_hummingbirds_lp.txt", head = T)

# Extract relevant posterior samples
sims_array <- fit_fm1$BUGSoutput$sims.array

# Keep only chains 4 to 6 (as identified from traceplots/diagnostics for good convergence)
sims_subset <- sims_array[, 4:6, ]

# Flatten to iterations*chains × parameters for easier processing
dim_combined <- dim(sims_subset)
n_iter_subset <- dim_combined[1]
n_chain_subset <- dim_combined[2]
n_param_subset <- dim_combined[3]

sims_flat <- matrix(NA, nrow = n_iter_subset * n_chain_subset, ncol = n_param_subset)
for (p in 1:n_param_subset) {
  sims_flat[, p] <- c(sims_subset[,,p])
}
colnames(sims_flat) <- dimnames(sims_subset)[[3]]
post_subset <- as.data.frame(sims_flat)

# Create prediction grid for hour
hour_seq <- seq(min(ObsCovs$hora_decimal, na.rm = TRUE),
                max(ObsCovs$hora_decimal, na.rm = TRUE), length.out = 100)
mean_flower <- mean(ObsCovs$flower_abun, na.rm = TRUE)

# Create prediction grid for flower abundance
flower_seq <- seq(min(ObsCovs$flower_abun, na.rm = TRUE),
                  max(ObsCovs$flower_abun, na.rm = TRUE), length.out = 100)
mean_hour <- mean(ObsCovs$hora_decimal, na.rm = TRUE)

# Predict detection probability across hour_seq
pred_hour <- sapply(1:nrow(post_subset), function(i) {
  logit_p <- post_subset$beta0[i] +
                  post_subset$beta_hour[i]*hour_seq +
                    post_subset$beta_flower_p[i]*mean_flower
  plogis(logit_p)
})

# Summarize posterior predictive mean and 95% credible interval for hour
df_hour <- data.frame(
  hour = hour_seq,
  p_mean = rowMeans(pred_hour),
  p_low = apply(pred_hour, 1, quantile, probs = 0.025),
  p_high = apply(pred_hour, 1, quantile, probs = 0.975)
)

# Predict detection probability across flower_seq
pred_flower <- sapply(1:nrow(post_subset), function(i) {
  logit_p <- post_subset$beta0[i] +
                 post_subset$beta_hour[i]*mean_hour +
                    post_subset$beta_flower_p[i]*flower_seq
  plogis(logit_p)
})

# Summarize posterior predictive mean and 95% credible interval for flower abundance
df_flower <- data.frame(
  flower_abun = flower_seq,
  p_mean = rowMeans(pred_flower),
  p_low = apply(pred_flower, 1, quantile, probs = 0.025),
  p_high = apply(pred_flower, 1, quantile, probs = 0.975)
)

library(ggplot2)

### Plot: Hour Effect on Detection
ggplot(df_hour, aes(x = hour, y = p_mean)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_ribbon(aes(ymin = p_low, ymax = p_high), alpha = 0.2, fill = "steelblue") +
  labs(y = "Detection probability", x = "Time of the day", title = "Effect of Hour on Detection") +
  theme_minimal()

### Plot: Flower Abundance Effect on Detection
ggplot(df_flower, aes(x = flower_abun, y = p_mean)) +
  geom_line(color = "darkgreen", size = 1) +
  geom_ribbon(aes(ymin = p_low, ymax = p_high), alpha = 0.2, fill = "darkgreen") +
  labs(y = "Detection probability", x = "Flower abundance", title = "Effect of Flower Abundance on Detection") +
  theme_minimal()

### Plot: Abundance across Seasons
# Convert subset_chains to a combined data frame
subset_df <- as.data.frame(do.call(rbind, subset_chains))

# Extract only the columns corresponding to N[i,t]
N_cols <- grep("^N\\[", colnames(subset_df), value = TRUE)
N_samples <- subset_df[, N_cols]  # this is [iterations x (sites × seasons)]

# Convert to array 
n_sites <- 10
n_seasons <- 4
n_iter <- nrow(N_samples)

N_array <- array(NA, dim = c(n_iter, n_sites, n_seasons))

for (s in 1:n_sites) {
  for (t in 1:n_seasons) {
    colname <- paste0("N[", s, ",", t, "]")
    N_array[, s, t] <- subset_df[[colname]]
  }
}

mean_N_per_season <- apply(N_array, c(1,3), mean)  # [iterations, seasons]

season_summary <- apply(mean_N_per_season, 2, function(x) {
  c(mean = mean(x), lower = quantile(x, 0.025), upper = quantile(x, 0.975))
})

season_df <- data.frame(
  Season = 1:ncol(mean_N_per_season),
  Mean = season_summary["mean", ],
  Lower = season_summary["lower.2.5%", ],
  Upper = season_summary["upper.97.5%", ]
)
season_df$Season <- factor(season_df$Season, labels = c("Spring", "Summer", "Autumn", "Winter"))

ggplot(season_df, aes(x = Season, y = Mean)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0) +
  labs(y = "Estimated Abundance", x = "Season") +
  theme_minimal()
```
